msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-12 paper.tex
Revisions.

The term Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a graphical
representation of the coalescent with recombination stochastic process. This
formulation was complementary to Hudson's earlier
work~\citep{hudson1983properties}, and substantially increased our theoretical
understanding of the recombination process. In Griffiths' ARG formulation, we
regard a realisation of the coalescent with recombination as a graph in which
nodes represent common ancestor or recombination events, and edges represent
lineages through which ancestral material flows. We may consider the ``big''
ARG, in which we wait an exponentially large time until all sampled lineages
coalescence into a single ancestor~\citep{ethier1990two}, or the ``little'' ARG
in which we wait until all marginal trees have fully coalescenced (like
Hudson's algorithm).

% Seems somewhat unfair to single out Iain and Alywyn here, but a systematic
% review would be long and tedious, and I'm sure they won't be too upset about it.
However, this original definition of the ARG in terms of a stochastic process
is rarely what is meant when the term is currently used. The term has come to
essentially mean a record of genetic ancestry as a graph; for example,
\cite{mathieson2020ancestry} informally define ARGs as the subset of an
individual's pedigree from which they have inherited ancestral material. Such
imprecision is unhelpful, since there are any number of graphical encodings of
genetic ancestry, and referring to the original Griffiths papers does little
to clarify. As we shall see, there are situations in which the precise form of the
graph is important, and conflating high-level representation of genetic
ancestry with this precise mathematical object is not helpful.
One may wonder if a more generic term such as ``genetic pedigree''
would be a more useful description, in such cases where the precise form
of the encoding is not important.

% ~\cite[.e.g.][]{minichiello2007mapping}
% \cite{minichiello2006mapping} are somewhat more explicit
% "An ARG provides a record of all coalescence and recombination events since the
% divergence of the sequences under study and specifies a complete genealogy at
% each genomic position
% \citep{rasmussen2014genome}
For our purposes here, an ARG is a realisation of the coalescent with
recombination, in the Griffiths sense.
As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
essentially works by dynamically generating and traversing a little ARG.
The graph is not explicitly represented in memory, but is partially
present through the extant lineages and the ancestral material they carry
over time. Since the ARG is a large and unwieldy object
(Fig.~\ref{fig-recombination-perf}B) we do not use it as output, but
rather store the information required to recover the genealogical
history as nodes and edges in a tree sequence (see section XX),
or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.
This approach is one of the key reasons for \msprime's efficiency.

Although the tree sequences output by \msprime\ contain full information
about the genealogical trees, their correlation structure along the genome,
and the ancestral haplotypes on which coalescences occured, information
is lost in this mapping down from ARG space. In particular, we lose
information about individuals who were common ancestors but where
no coalescences occured and also information about the precise time
and genomic location of recombination events. In most cases, such
information is of little relevance as it is in principle unknowable,
but there are occasions in which is useful. We therefore provide
the \texttt{record\_full\_arg} option in \msprime\
to store a representation of the compete ARG traversed during simulation.
This is done by storing extra nodes (marked with specific flags, so they
can be easily identified later) in the tree sequence.
Fig.~\ref{fig-recombination-perf}B shows that these extra ARG nodes
and edges quickly come to dominate the output as we increase the sequence length,
and the resulting tree sequence files are many times larger
than the standard minimal tree sequence representation.

@@

% \begin{comment}
% Previous draft of the ARG feature retained here as a comment.

% The definition of ARGs is vague and confusing.
% \cite{minichiello2006mapping} define it as the data structure
% rather than the original stochastic
% process~\citep{griffiths1991two,griffiths1997ancestral}.
% \cite{rasmussen2014genome} have a slightly different approach.
% Sometimes the ARG is useful.
% The tree sequence output for \msprime\ can be turned into
% an ARG using the \texttt{record\_full\_arg} option, where we store
% recombination nodes.

% We also provide an interface to compute the likelihood of a given ARG
% under the~\cite{kuhner2000maximum} model.
% \end{comment}

>>>>

The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a graphical
representation of the coalescent with recombination stochastic process. This
formulation was complementary to Hudson's earlier
work~\citep{hudson1983properties}, and substantially increased our theoretical
understanding of the recombination. In Griffiths' ARG formulation, we
regard a realisation of the coalescent with recombination as a graph in which
nodes represent common ancestor or recombination events, and edges represent
lineages through which ancestral material flows. We may consider the ``big''
ARG, in which we wait an exponentially large time until all sampled lineages
coalescence into a single ancestor~\citep{ethier1990two}, or the ``little'' ARG
in which we wait until all marginal trees have fully coalescenced (like
Hudson's algorithm).

% Seems somewhat unfair to single out Iain and Aylwyn here, but a systematic
% review would be long and tedious, and I'm sure they won't be too upset about
% it.
However, this original definition of the ARG in terms of a stochastic process
is rarely what is meant when the term is currently used. The term has come to
essentially mean a representation of genetic ancestry as a graph; for example,
\cite{mathieson2020ancestry} informally define an ARG as the subset of an
individual's pedigree from which they have inherited ancestral material. Such
imprecision is unfortunate, since there are any number of graphical encodings of
genetic ancestry, and referring to the original Griffiths papers does little to
clarify. As we shall see, there are situations in which the precise information
stored of the graph is important, and conflating a high-level
discussion of genetic ancestry with this mathematical object is not
helpful. One may wonder if a more generic term such as ``genetic pedigree''
would be a more useful description, in such cases where the precise form of the
encoding is not important.

For our purposes here, an ARG is a realisation of the coalescent with
recombination, in the Griffiths sense.
As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
works by dynamically generating and traversing a little ARG.
The graph is not explicitly represented in memory, but is partially
present through the extant lineages and the ancestral material they carry
over time. Since the ARG is a large and unwieldy object
(Fig.~\ref{fig-recombination-perf}B) we do not use it as output, but
rather store the information required to recover the genealogical
history as nodes and edges in a tree sequence (see section XX),
or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.

Although the tree sequences output by \msprime\ contain full information
about the genealogical trees, their correlation structure along the genome,
and the ancestral haplotypes on which coalescences occured, information
is lost in this mapping down from ARG space. In particular, we lose
information about ancestral haplotypes that were common ancestors but
in which no coalescences occured, and also information about the precise time
and genomic location of recombination events. In most cases, such
information is of little relevance as it is in principle unknowable,
but there are occasions such as visualisation or computing likelihoods (see
below) in which is useful. We therefore provide
the \texttt{record\_full\_arg} option in \msprime\
to store a representation of the compete ARG traversed during simulation.
This is done by storing extra nodes (marked with specific flags, so they
can be easily identified later) and edges in the tree sequence.
Fig.~\ref{fig-recombination-perf}B shows that this extra ARG information
quickly comes to dominate the output as we increase the sequence length,
and the resulting tree sequence files are many times larger
than the standard minimal tree sequence representation.

@@
